---
title: 'Lab 2: Data wrangling'
output: html_document
---
In this first "real" lab we'll go over one of the most important skills in data science: Data Wrangling. It's the art of cleaning and manipulating our data to prepare it for our analyses and visualizations. A common estimate is that for most machine-learning projects, 80% of the time is spent on data wrangling, and the remaining 20% is spent on actually setting up and training the machine learning model.

Over the next two weeks, we'll do a lot of similar tasks, just in different ways. Today, we will focus primarily on what is called Base R, which means using functions that are built into R. Next week, in class, we'll see a collection of packages called `tidyverse`. The `tidyverse` contains tons of extremely useful functions that make data cleaning and manipulation a breeze. You might wonder then, why we aren't learning that right away. In R, you'll learn multiple ways to do things. Some times you'll read code that is written in a different way than you'd write it but it does the same thing, and it's good practice to get comfortable with the fact there are often multiple ways to do the same thing in R. 


We'll start with basic data manipulation skills. It is useful to FIRST try to visualize in our head what we want to do with the data,and THEN look for the code that can do that for us. Once you learn some useful functions, you will be able to combine them to manipulate the data to your will like a wizard.


# Part 1: Base R

### 1. Getting an overview of the data
First we'll explore different ways to get an overview of the dataset.
Let's use the gapminder dataset, available in the `dslabs` package. Because the data is available in the package, we can load it by simply using `data(gapminder)`. First we'll look at the dataset to see what it contains. We will use `summary()`, `head()` and `str()`.

```{r, message=FALSE}
# install.packages('dslabs') ### Only run if you haven't installed dslabs
library(dslabs)
data(gapminder)
df <- gapminder
```

##### 1) What type of information do each one of these functions provide about our dataset?

##### 2) How is the dataset ordered?

##### 3) What range of years does our dataset contain?


```{r}
summary(df)

head(df)

str(df)
```



### 2. Ordering and subsetting our dataset
We can subset our dataset in different ways. For example, the brute-force approach: we can select the cell in row 1, column 3 using `gapminder[1,3]`. We can also select a range of rows using `:`; i.e. we can select rows 2 to 5, column 1 using `gapminder[2:5,1]`.

```{r}
### Subsetting via numeric index

gapminder[1,3] ### Select row 1, column 3

gapminder[2:5,1] ### Select rows 2-5, column 3

gapminder[c(1,3,5), c(1,3,5)] ### Select rows 1,3,5 columns 1,3,5

gapminder[1,] ### entire first row, all columns (note we still need comma)

gapminder[,1] ### entire first column, all rows (note we still need comma)
```


##### 4) In general, selecting rows and columns via numeric indices may not always me the best idea. What are some disadvantages you can think of?


Another way to select columns is with a vector of column names

```{r}
### Select a subset of columns w/ a character vector
gapminder[, c('country', 'year', 'infant_mortality')]
```


__Filtering rows w/ a logical__:

We will see soon in class how to index rows w/ logical conditions. A logical is just a vector of true and falses. Let's look first at a simply example of a vector, before moving to our full dataset.

```{r}
### Save out the numbers 1-10
a <- 1:10

### Logical comparisons
a < 3

a <= 2 | a > 8

a >= 3 & a <= 7

### Subsetting a via logical comparisons
### Just returns the elements that are true
a[a < 3]
a[a <= 2 | a > 8]
a[a >= 3 & a <= 7]

```


```{r}
### Filter to just the united states
gapminder[gapminder$country == 'United States',]

### Combine Filter + Subset
gapminder[gapminder$country == 'United States', c('country', 'year', 'infant_mortality')]
```

What happens w/ NAs? We see that infant mortality has `NA` values

```{r}
gapminder[gapminder$infant_mortality > 10,]
```

When we do this, we see that values where infant mortality are `NA` are kept. We can get around this by using `which()`. Which returned the numeric index of values that are `TRUE` in a list.

The `which()` function returns the index of rows in a data frame that meet a certain criteria. For example, if we want to know the index of the rows for Algeria, we can type the following:

```{r}
which(gapminder$country == "Algeria")
```

We can use this to select rows within our data frame by using brackets `[ ]` and placing our `which()` function where we would place the row number.
```{r}
head(gapminder[which(gapminder$country == "Algeria"), ])

head(gapminder[which(gapminder$country == "Algeria"), "infant_mortality"])
```

Now, let's use this to filter rows w/out NA values.
```{r}
ix <- which(gapminder$infant_mortality > 10)
ix
gapminder[ix,]
```


##### 5) Create a data frame called `algeriagdp` that includes only year and GDP data for Algeria.



### 3. Ordering Data

The function order returns the numeric indices needed to sort the data
```{r}
a <- c(1,3,5,2,4,8)
b <- c('apple', 'orange', 'banana', 'pear', 'grape', 'lemon')

### Sort a
sort(a)

order(a) ### order in which to get the elements of a so that it is sorted
a[order(a)] ### equivalent to sort
b[order(a)] ### sort b by a


### application to the data frame
gapminder[order(gapminder$infant_mortality),]
gapminder[order(gapminder$life_expectancy, decreasing = T),]

```

##### 6) Create a data frame called `df2` that is ordered so that it shows the data for each country (all years), then the next country (all years), etc.


### 4. Creating new data

Lastly, we will look at creating new data or editing data.

```{r}
### Make a new column
### GDP/Population = gdp per capita
gapminder$gdp_per_capita <- gapminder$gdp/gapminder$population

### Make contry a character (not factor) and then rename United States to USA
gapminder$country_chr <- as.character(gapminder$country)
gapminder$country_chr[gapminder$country_chr == 'United States'] <- 'USA'
```


# Part 2: Tidyverse
Now, we will do the same types of operations but using the tidyverse functions.

We can use the `filter()` function to select rows according to values of a column, and we can use `select()` to select only specific columns according to their name. If we want a specific set of rows, we can use `slice()`.

```{r, warning = F, message = F}
library(dplyr)
```

First, let's see filter. Notice that we don't need to use `df$` infront of the names, after we supply df as the first argument to the function.
```{r}
### filter
filter(df, country == 'United States')

### Notice that filter automatically throws out NA -- no need for which()!
filter(df, infant_mortality < 10)

### Multiple combinations
filter(df, infant_mortality < 10, year > 1990)

```

#### 1) How many rows have `infant_mortality > 20` and `life_expectancy > 70`?


Next we can use select to get column names. select doesn't require that we uses quotes around names.
```{r}
### Select 3 columns
select(df, country, year, infant_mortality)

### Select everything except the region column
select(df, -region)
```

```{r}
### 1st row of df
slice(df, 1)

### rows 20-50 
slice(df, 20:50)
```


We can also order a data frame using `arrange()`


```{r}
### Default is increasing
arrange(df, infant_mortality)

### use desc() for decreasing
arrange(df, desc(infant_mortality))

### sort by continent, then region
arrange(df, continent, region)
```

##### 2) Create a dataset that includes country, year, gdp and population for countries in South-Eastern Asia between 1965 and 2010. Order the data frame by country, then year. Confirm that your code is doing what it's supposed to.


### 2. The pipe operator %>%

To generate the result above, you may have needed steps in between. We can use the pipe operator `%>%` to avoid using intermediate steps. This makes for much more intuitive and fun data wrangling.

For example, we can create the data frame mentioned above in an elegant, intuitive way:
```{r}
algeriagdp <- 
  gapminder %>% 
  filter(country == 'Algeria') %>% 
  select(gdp,year)
```

##### 3) Now, using the pipe operator, create a dataset that includes country, year, gdp and population for countries in South-Eastern Asia between 1965 and 2010. Order the data frame by country, then year. Try to do it all in one command, starting from the `gapminder` dataset, using the pipe operator. Confirm that your code is doing what it's supposed to.

```{r}
# your code here
```

### 4. Creating variables

We can create variables using `mutate()`.

```{r}

```

##### 1) Create a variable with the approximate number of women in the population. Assume that 52% of the population are women.

##### 2) Calculate an approximation of the number of children born each year based on the `fertility` column, which displays the expected number of children per woman over their reproductive life.
For this, assume an average reproductive life of 25 years.

##### 3) Look up and explore the function `lag()`. Use it to calculate the change in population from the year prior, and assign it to a column called `popchange`.

```{r}

```


### 5. Summary

As we saw in class, the function `summarize()` provides a useful way to combine data from different rows.

##### 1) Create a dataset with the total world population by year.

##### 2) Create a dataset with the gdp per capita by world region for the year 2010. Be careful of the order when using the pipe operator `%>%`

##### 3) Calculate the increase in gdp per capita from the year prior by region and year.


```{r}
#1 

#2 

#3 

```


You can find a list of our spells... ehhhm, functions [here](https://rstudio.github.io/cheatsheets/data-transformation.pdf).


And a list of other useful spellbooks (cheat sheets) [here](https://posit.co/resources/cheatsheets/). 

