---
title: "instructor recap notes - lec 17"
date: "2023-10-23"
output: html_document
---


### 0. Random quiz: 
How to check the first few rows of your data?

```{r random quiz and setup, message=FALSE}
# load the required libraries
library(tidyverse)
library(ggplot2)
library(dslabs)

# read in the example data file
filename <- "https://raw.githubusercontent.com/rafalab/dslabs/master/inst/extdata/fertility-two-countries-example.csv"
wide_data <- read_csv(filename)

# check the first few rows of your data
head(wide_data)
```

### 1. Reshaping data with `pivot_longer()` and `pivot_wider()`
Both functions are from the library `tidyverse` which we already loaded in the previous code chunk. 

#### 1.0  Concept of data shapes

- **Wide data**: Observations have many columns, often with different variables. This is often the format in which data is collected or presented in spreadsheets.
  
- **Long data**: Observations are in a single column, with another column specifying the variable they measure. This is often required for data visualization or modeling functions in R.

#### 1.1 The `pivot_longer()` Function:
Imagine you have data from a school where test scores for Midterm and Final for each student are in separate columns. What if you wanted to analyze scores across subjects, or visualize them? You'd want these scores in a 'long' format.

**Example Data**:
```{r pivot longer example}
library(tibble)
data_wide <- tribble(
  ~Student, ~HW1, ~HW2, ~HW3,
  "John", 90, 95, 85,
  "Jane", 95, 90, 88
)

# # Equivalent to 
# data_wide <- data.frame(
#   Student = c("John", "Jane"),
#   HW1 = c(90, 95),
#   HW2 = c(95, 90),
#   HW3 = c(85, 88)
# )

print(data_wide)
```

**Transformation**:
```{r transform to longer}
library(tidyr)
data_long <- data_wide %>% 
  pivot_longer(cols = c(HW1, HW2, HW3), names_to = "Homework", values_to = "Score")
print(data_long)
```


#### 1.2 The `pivot_wider()` Function:
Now, let's consider the opposite situation. You have data in a long format where each row represents a student's score in a specific subject. What if you want to compare scores in different subjects side by side? You'd want these in a 'wide' format.
**Transformation**:
```{r transformation}
data_wide_again <- data_long %>% 
  pivot_wider(names_from = Homework, values_from = Score)
print(data_wide_again)
```


#### 1.3 Conclusion:
- It's essential for data visualization in tools like `ggplot2`.
- Many statistical functions expect data in a long format.
- Pivoting wider can help in comparing multiple variables side by side.

### 2. Today's lecture
- HW3 released; Due Nov 5 (2 late days)
- Midterm after HW3; 2 weeks
- Final project options released on our [course website](https://coredatascience-fa23.github.io/project/). Follow the instructions and fill out the [preference form](https://docs.google.com/forms/d/e/1FAIpQLSfvhegiqPG-jQtQvqaIz6jnyLrSv-sOpiF4V3qvQdSTMHK3zQ/viewform) by Nov 6. 
- Today: date and times; advanced data wrangling practice
- Lab this week: more practices on advanced data wrangling

