---
title: "instructor recap notes - lec 18"
date: "2023-10-25"
output: html_document
---


### 0. Random quiz: 
1. How to check the usage of a function you've never seen / not familiar with? 
2. How to summarize a variable by another factor variable? (e.g. summarize the mean of grades by homework number)

```{r random quiz and setup, message=FALSE}
# load the required libraries
library(tidyverse)
library(ggplot2)
library(dslabs)
library(lubridate)

?mdy
?lubridate::mdy
```
### 1. Joining tables 
Joining tables is a fundamental operation in data processing and analysis, allowing you to combine information from multiple tables based on common attributes. In R, the `dplyr` package provides various functions to perform different types of joins. Here are six commonly used join functions. 

1. **Inner Join (`inner_join`):**
   - Combines matching rows from two tables based on a common column (key).
   - Only rows with matching keys in both tables are included in the result.
   
```{r inner join}
# Example data
df1 <- data.frame(ID = c(1, 2, 3), Name = c("Alice", "Bob", "Charlie"))
df2 <- data.frame(ID = c(2, 3, 4), Age = c(25, 30, 35))

# Inner Join
inner_join(df1, df2, by = "ID")
# With pipe:
df1 %>% inner_join(df2, by = "ID")
```

2. **Left Join (`left_join`):**
   - Combines all rows from the left table and matching rows from the right table.
   - Non-matching rows from the right table will have NA values.
   
```{r left join}
# Left Join
left_join(df1, df2, by = "ID")
# With pipe:
df1 %>% left_join(df2, by = "ID")
```

3. **Right Join (`right_join`):**
   - Combines all rows from the right table and matching rows from the left table.
   - Non-matching rows from the left table will have NA values.
   
```{r right join}
# Right Join
right_join(df1, df2, by = "ID")
# With pipe:

```

4. **Full Join (`full_join`):**
   - Combines all rows from both tables.
   - Non-matching rows will have NA values in the columns from the table where there is no match.
   
```{r full join}
# Full Join
full_join(df1, df2, by = "ID")
# With pipe:

```

5. **Semi Join (`semi_join`):**
   - Keeps all rows from the left table that have matching keys in the right table.
   - Does not include any columns from the right table.
   
```{r semi join}
# Semi Join
semi_join(df1, df2, by = "ID")
# With pipe:
```

6. **Anti Join (`anti_join`):**
   - Keeps all rows from the left table that do not have matching keys in the right table.
   - Does not include any columns from the right table.
   
```{r anti join}
# Anti Join
anti_join(df1, df2, by = "ID")
# With pipe:

```

In each of these examples, the `by` argument specifies the column(s) on which to perform the join. **It's important to ensure that the join keys have matching data types in both tables to avoid unexpected results.**


### 2. Date and Times in R 
Working with dates and times can be tricky, but the `lubridate` package in R simplifies the task. 
The `lubridate` package provides tools to work with date-time data. It simplifies the process of parsing, manipulating, and doing arithmetic with dates and times in R. 


#### 2.1 Parsing dates and times from strings
`lubridate` provides functions to parse dates and times from strings.

1. **ymd(), dmy(), mdy(), ...**
   - Parse dates from strings.
```{r parse dates}
   ymd("2023-10-25")  # Year-Month-Day
   dmy("25-10-2023")  # Day-Month-Year
   mdy("10-25-2023")  # Month-Day-Year
```

2. **ymd_hms(), dmy_hms(), mdy_hms(), ...**
   - Parse dates and times from strings.
```{r parse dates and times}
   ymd_hms("2023-10-25 12:34:56", tz = "EST")
```

#### 2.2 Manipulating date-time data

1. **year(), month(), day(), ...**
   - Extract components of a date-time object.
```{r manipuate date time}
   date <- ymd("2023-10-25")
   year(date)  # 2023; numeric
   month(date)  # 10; numeric
   day(date)  # 24; numeric
```

2. **today(), now()**
   - Get today's date and the current date-time.
```{r today now}
   today()
   now()
   ymd("2023-12-11") - today()
```

#### 2.3 Arithmetic with date-time data

1. **Duration:**
   - Create a duration object, which represents a time span.
```{r duration}
   dur <- ddays(1) + dhours(3)  # 1 day and 3 hours
```

2. **Period:**
   - Create a period object, which represents a time span in human terms (days, months, etc.).
```{r period}
   per <- days(1) + months(1)  # 1 month and 1 day
```

3. **Date-time arithmetic:**
   - Add or subtract duration or period objects from date-time objects.
```{r date time arithmetic}
   date + dur  # add duration
   date + per  # add period
```

#### 2.3 Plot
# Create a dataset
```{r plot time}
data <- data.frame(
  Date = seq(as.Date("2023-01-01"), by = "month", length.out = 12),
  Value = c(10, 20, 15, 25, 30, 35, 40, 45, 50, 55, 60, 65)
)

# Create the plot
data %>% ggplot(aes(x = Date, y = Value)) +
  geom_line() +  # Line plot
  geom_point() +  # Points
  labs(
    title = "Monthly Values in 2023",
    x = "Date",
    y = "Value"
  ) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b %Y") +  # Format x-axis dates
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + # Adjust x ticks orientation 
  labs(x = "date",
       y = "value")

```

### 3. Today's lecture
- HW3 released; Due Nov 5 (2 late days)
- Midterm to be released on Nov 8 (due Nov 22) **no free late days**
- Final project options released on our [course website](https://coredatascience-fa23.github.io/project/). Follow the instructions and fill out the [preference form](https://docs.google.com/forms/d/e/1FAIpQLSfvhegiqPG-jQtQvqaIz6jnyLrSv-sOpiF4V3qvQdSTMHK3zQ/viewform) by **Nov 6**.
- Find a time (office hours preferably) to meet with TF (20 minutes) before Nov 17 to review your project. 
- Today: New module: probability and regression (yay!)
- Lab this Friday: more practices on advanced data wrangling

